Case-based Arabic Morphological Analysis

### 2010-12-07 tex/rp.tex
good shape

We think that the ability to extract such orders
between entities in documents can

@@

%\begin{figure}[tb]
\begin{figure}
\center{
\resizebox{.8\columnwidth}{!}
{\includegraphics{figs/detailed_submachine.pdf}
\label{f:chain}
}
\caption{CATARM detailed diagram for the chain of narrators extractor}
}
\end{figure}

@@

The difficulty of applying text mining techniques to Arabic text 
documents lays in the absence of automated tools that understand 
the unique features of the Arabic language.
%We will make use of an example to illustrate the process.
Given an Arabic text document we want a tool that can segment 
and isolate all names, dates, time of the day occurrences,
tool names, and locations as entities.
Then we want another tool to identify relations amongst 
these entities such as identities and orders.
An order between two entities can be simply 
their order of appearance in the document, or their
numerical 
 We want another tool to look at the verbs and actions in the sentences and try to build relations between the identified entities based on these verbs.
 If we were successful to do that,
 we now want to order all entities associated with dates based on a chronological order.

To do the above,
 we need tools that are able to automatically differentiate between names,
 verbs,
 adjectives and other vocabulary structures.
 This is not a simple task with the Arabic language.
 For example,
 Arabic grammar allows name-based sentences or sentences with no 
    verbs \noTrNoVocRL{^gml ismiyT} .
 Another unique feature is the possibility to have names in Arabic 
 that express action such as noun verbs 
 \noTrNoVocRL{ism f`l, f-a`l $\ldots$}.

 In addition to these structural differences,
 there are features in the Arabic language that need special 
 treatment such as derivatives and stems 
 \noTrNoVocRL{alm^staq-at w al^g_dwr}.
 We also need to build dictionary tools that associate words and phrases based on their meanings and their context.
 Often time,
 and depending on the application and the user of the tool,
 we need to allow the user to assign meanings and semantics for patterns in the text.

In the following we discuss our research methodology to develop the proposed Arabic text analysis tools.

Lexical analysis library: We need a library of routines that enables primitive lexical analysis of Arabic text.
 For example,
 simple primitive tasks such as comparing two strings need special attention in Arabic since each Arabic letter has four different forms depending on its position in the word; isolated form,
 start,
 middle and end of a word.
 Arabic diacritics should also be ignored most of the time when comparing two strings.
 These lexical analysis routines should also pay attention to the form of a letter when used in tokenization as a letter in and end of word form denotes the end of the word even if it was not followed with a space.
 Such a library can benefit from existing Unicode enabled string libraries such as Qt.
 It also can provide a layer of abstraction that hides the details of the encoding of Arabic letters used in the text documents.

Syntax analysis library: This library should allow sentence structure understanding.
 We will explore the best way to develop this library.
 Several options exist such as pattern based matching,
 tokenization and dictionary based lookups,
 or a language grammar that allows more than one matching rule.
 This library needs to be extendible to allow additions of new rules or patterns.

 (Note the ElixirFM does this)
 We will explore Arabic grammar books such as The Principles of Arabic~\cite{Sha73} and others~\cite{Abd00,Abd001}
 to augment current morphological analyzers~\cite{Tim04} and syntax analysis tools~\cite{Col09}.
 We will use relaxed NLP techniques to overapproximate results.
 Then and after a relational query,
 we will apply statistical analysis techniques to rank the matching results.

Dynamic modern Arabic dictionary: The dictionary will allow identities based on semantics,
 derivation rules,
 as well as user assigned rules.
 The dictionary will return modern grammatical categories for words classifying them as nouns,
 verbs,
 and adjectives instead of the classical Arabic categories that specify how the word is derived from an original radix.
 It remains to be explored how to represent these categories without exploding the size of the dictionaries with redundant words of multiple categories.
 This work will augment existing open source Arabic dictionaries and will allow for better morphological analysis and entity extraction.

An entity extraction library: We will explore building a library that allows extraction of entities from Arabic text.
 These entities can be proper names,
 dates,
 locations,
 and qualification adjectives to list few possibilities.
 The extractor will use the dynamic modern Arabic dictionary as well as the syntax library to extract entities from a sentence and to establish identity between entities across sentences.

A relation extraction library: We will explore building a library that allows extraction of relations from Arabic text.
 We will explore the use of pairing techniques to establish relations between the extracted entities.
 The relations will form graphs where the nodes are the entities and the previously extracted relations and the edges are the newly discovered relations.
 The routines in this library will take as input a template for an expected relation and will return a graph that describes whether this relation exists between the subject documents and where.
 These relations will be ranked using statistical analysis.
 The statistical criterion and the ranking methods remain to be decided.

A logic solver: This is a logical solver that will solve query expressions to resolve matching text and relations.
 We will explore the proposal of a logical query language that is restricted enough to be computable and expressive enough to allow interesting properties.
 We will use of the shelf solvers and matches from the entity and relation extractors to evaluate the expression.
 The choice of solvers remains to be decided after the query language is formed.

Tagger: We will explore the development of an automatic tagging interface that allows fast and guided inspection and correction mechanism.

We will integrate all the mentioned techniques in a scalable framework for Arabic text analysis.
 The framework as a whole will allow for exhausting Arabic text analysis and allow for extensions of the framework.
 A user should be able to iteratively and repetitively apply the abode techniques on his input text.

In the development cycle, we will use an Extreme programming methodology in the process of building the Arabic text mining framework.
 We will use open source tools, libraries, and frameworks to build our libraries.

Our research will concentrate on NLP techniques native to the Arabic language to analyze the Arabic text and we will avoid preprocessing that feeds common text and data mining techniques.
 We will build on existing open source programming and text mining tools as well as open source Arabic text analysis tools based on NLP such as MADA+TOKAN~\cite{Rot08}, NEMLAR~\cite{RAl09}, and IJAES~\cite{Int09}.

@@

@@

We anticipate that the research will lead to the production of an Arabic text analysis framework that allows relational queries between Arabic text documents.
 The components of the framework are listed in Section "Research Design and Methods" and are repeated for convenience as follows.
 \begin{enumerate}
 \item Lexical analysis library 
 \item Syntax analysis library
\item Dynamic modern Arabic dictionary
\item An entity extraction library
\item A relation extraction library
\item A logic solver 
\item Tagger
\end{enumerate}

We will use these tools to handle several case studies from different fields.
 In the following we discuss three case studies that we anticipate our proposed framework to solve.

{\bf The literature case study}
Assume we have two sets of texts.
 The first set is several books of historical accounts where each account is preceded by a chain of narrators who narrated the account.
 The second set is several books of biographies where each biography of a person may include an evaluation of the authenticity of that person.
 We want a tool that can answer a complex query such as relating all historical accounts in the first set that discuss women and vehicles (such as tamed animals) to the narrators of the accounts, and then relate these narrators to the second set of books of biographies with an emphasis on authenticity evaluations.

We need tools that can search and index the historical accounts to return a subset that relates to women and travel.
 Then we need tools that can identify names, and succession relations between names to build a chain of narrators for each historical account in the subset.
 We need tools that can identify the names in the biographies and detect evaluation adjectives.

Provided we have these tools we can exhaustively at the click of a button visualize the authenticity of all the historical accounts that may allow or forbid women from driving according to Islamic laws.
 We can go one step further to detect locations and dates and relations based on chronological orders and further question the consistency of the historical accounts.
 Such query takes currently huge amounts of manual effort that is also prone for error and far from being exhaustive.

We here envision that text mining applied to the authenticity of historical accounts will have a huge impact in the field.

Note that most of the Islamic history and Hadith books as well as a huge library of Arabic literature books are already digitized and available on the internet.

{\bf The security case study}

Given a huge set of investigation reports, an officer may throw a list of names of convicted people and check all the reports for possible relations between them in terms of criminal action, locality, or third party names unknown to the officer who happen to know the convicts.
 The query may return at the click of a button, two convicted people who had no direct relation in the reports, but who are related only through a third person unknown to the investigator.
 Such a query is very hard to answer manually.

{\bf The health case study}

Given a set of clinical reports gathered by the Order of Medical Doctors, the ministry of health may query to relate all reports on a specific sickness to the doctors treating it and evaluate the response to the different types of medications as classified in different areas in the country each with a different climate.
 Such query, while possible at a click of a button with text mining tools if the data is available in digital form, would take a huge manual effort to conduct surveys, collect and gather the data.
 In this case, digitizing the data may be a less expensive effort.

>>>>

@@

@@

%The difficulty of applying text mining techniques to Arabic text 
%documents lays in the absence of automated tools that understand 
%the unique features of the Arabic language.
%We will make use of an example to illustrate the process.
Given an Arabic text document we want a tool that can segment 
and isolate all names, dates, time of the day occurrences,
tool names, and locations as entities.
Then we want another tool to identify relations amongst 
these entities such as identities and orders.
We want another tool to look at the verbs and actions in 
the sentences and try to build relations between the identified 
entities based on these verbs.
If we were successful to do that,
we now want to order all entities associated with dates based 
on a chronological order.

An order between two entities can depend on 
their simple order of appearance in the document, their 
numerical values in case they happened to be numbers, or
the distance between them in case they were cities. 
The relations can also be more sophisticated and 
based on templates provided by users or by 
models extracted from training data sets.

To do the above, 
we need tools that automatically differentiate between 
names, verbs, adjectives and other structures.
This is not a simple task with the Arabic language.
For example, Arabic grammar allows name-based sentences 
or sentences with no verbs \noTrNoVocRL{^gml ismiyT} .
Another unique feature is the possibility to have names in Arabic 
that express action such as noun verbs 
\noTrNoVocRL{ism f`l, f-a`l $\ldots$}.

In addition to these structural differences,
there are features in the Arabic language that need special 
treatment such as derivatives and stems 
\noTrNoVocRL{alm^staq-at w al^g_dwr}.
Simple primitive tasks such as comparing two strings need 
special attention in Arabic since each Arabic letter has 
four different forms depending on its position in the word; 
isolated form, start, middle and end of a word.
Arabic diacritics should also be ignored most of the time 
when comparing two strings.
We also need to build dictionary tools that associate words 
and phrases based on their meanings and their context.
Often time,
and depending on the application and the user of the tool,
we need to allow the user to assign meanings and semantics 
for patterns in the text.

\subsection{Literature authentication case study}
\label{s:design:lit}

The diagram in Figure~\ref{f:literature} describes our 
high level vision for what the literature case study
will look like in the CATARM framework. 
The boxes represent 
computational blocks provided by CATARM, 
ovals represent data blocks which are mainly the 
inputs and the outputs of the computational blocks. 
Dotted data blocks represent intermediate results,
and solid and shaded blocks represent the initial input 
and the final output respectively.

The computational components can be direct CATARM components
or custom user components built from other CATARM components.
We feed the hadith books to a user Narrator Chain Extractor
component that we will discuss later in 
Section~\ref{s:design:chainext}. We  obtain a set of chains
of narrators that we feed to a morphological analyzer (MA) 
component controlled by a merger controller. 
The merger controller will look at the morphological analysis
of the narrator names and will decide to merge two narrators in 
one graph node in case they were morphologicaly equal. 
It will also give feedback to the MA component to stop the
analysis when a it reaches a satisfactory result.
We consider this graph as a partial order relation between 
narrators where a narrator precedes another narrator if 
he narrated from him.

The biography books are fed to another MA unit controlled by 
a name and qualification pair controller that generates a 
map consisting of pairs of narrator names associated with 
qualifications. 
Then the graph of narrators and the narrator qualification map
are both processed through a locator component that computes
a graph coloring routine to color the graph of narrators. 
The color changes whener two consecutive narrators are far away
from each other in the text. 
The locator then looks at connected subgraphs of the same color,
and finds a median node in each of them to be the narrator 
associated with the qualifiers of the color. 
Then the locator also annotates the narrator with the 
corresponding qualifications. 
The result is an annotated partial order graph of narrators
with authentication qualification annotations.

\subsection{Security case study}
\label{s:design:sec}

The diagram in Figure~\ref{f:security} describes our 
high level vision for what the solution of the 
security case study
will look like in the CATARM framework. 
The inputs are two separate sets of documents. 
The first set includes crime scene investigation reports, 
and the second set includes a 
interrogation reports with crime suspects as well as 
a database of previous interrogation reports. 
The CATARM system also takes as input a database of geographical
locations and a database of personal records. 
The aim is to check for inconsistencies in the iterrogation
reports and to identify possible connections between the suspects
and the crime scene.

An MA with date, place, and name extraction controller extracts
names, dates and places from the investigation reports. 
We pass the dates to a sorting component to produce $G_D$ 
a partial order graph between the dates. 
We look up the names in the personal records database and find out
$G_N$ a family dependency graph between the names. 
We pass the places to a mapper that builds another $G_G$,
a graph relating the places using a geographical database.

We pass the interrogation reports to an MA with a relation 
extraction controller that builds $G_R$. 
The graph $G_R$ relates the interrogated suspects
to each other, to places, and to dates based on a simple
neighborhood threshold computed using a statistical feature.

We feed $G_D, G_F, G_G,$ and $G_R$   
to a consistency checker.
The consistency checker considers the three dimensions 
$N$ names, $G$ locations, and $D$ dates and projects $G_R$
over each one of them.
For each dimension $k$, it computes a local measure 
$\alpha_k$ of how much $G_R|_k$ simulates $G_k$. 
It flags the components in the graph where the accumelated 
measure $\alpha= \Sigma_{k\in\{D,N,G\}} \alpha_k$ is
below a threshold $\theta$ as inconsistent.

\subsection{CATRAM components}
\label{s:design:components}

To do the above we need CATARM to provide us with several 
computational components such as a morphological analyzer, 
statistical tools, graph algorithms, and other computational
components. 
In the following we discuss the components needed for 
CATARM and we propose a methodology to 
use and extend existing tools, and develop new tools to
build the components of CATARM.

\subsection{Case-based morphological analyzer}

Morphological analyzers for the Arabic language
such as Buckwalter~\cite{Buckwalter:02},
Beesly~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
and ElixirFM~\cite{Otakar:07} take an Arabic word as input
and produce several valid morpholgical solution for the word. 
Each solution is a sequence of morphemes that form the word.
Each morpheme is associated with several tags such as 
morphological, part of speech, grammar, and semantic tags. 
These analyzers are essential in several text analysis tools such 
as spell checkers as well as NLP frameworks~\cite{Col09}.
Other morphological analyzers such as 
Amira~\cite{Diab:07,Benajiba:07},
MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
use machine learning and support vector machines (SVM) 
to enhance the accuracy at the expense of efficiency.

The current use of morphological analyzers suffers from 
several problems. 
Analyzers that enumerate all possible partitions of a word
and match them against concatenative rules are computationaly 
expensive. 
Analyzers that exhaustively enumerate all possible solutions 
for a word may hurt performance and may
not be necessary or appropriate in some case studies~\cite{Maamouri:10}. 
For example, the first solution may be the one needed and 
we may not need to list the rest of the solutions. 
A white space delimited token may have more than one word
as is the case with ``run-on words''.

We propose to make use of the current existing morphological
analyzers to build a flexible and more efficient
morphological analyzer. 
The analyzer will be an essential
component in the CATARM framework. 
It will be a case-based analyzer as it will make guided decisions
based on feedback from the NLP case study and it will prune
invalid solutions on the fly.

Case-based analysis can guide the morphological analysis
to prune lots of the solutions and to bias the order of considered
decisions. 
For example, in the hadith literature case study pesented above, 
amongst other things, we are intersted in checking whether a 
word is a name.  
We can bias the order of considered decisions to consider the 
morphemes that can be names first. 
If one of them matches, we can ignore the rest of the solutions. 
In case a prefix is decided to be of a morphological category
that does not connect with a name, then we can just ignore
the analysis of the rest of the word. 
This is more efficient than generating all possible partitionings
of the word, checking for morphological correctness of each 
partition, and then inspecting the part of speech tags of each
correct solution. 
We desire to build a morphological analyzer that allows a 
case-based controller to intervene at every single decision
without much overhead.

\subsection{CATARM Query language}
\label{s:design:query}

We propose the CATARM query language with a user friendly
visual representation similar to that in Figures~\ref{f:literature} and
~\ref{f:security}. 
The basic elements in the query language will be entities, relations
and tags similar to those reported by a morphological analyzer. 
The query language will also have connectives such as set operations, 
relational operations, 
scalar and Boolean arithmetic operations, 
statistical computations, graph algorithms, and text algorithms
form terms and predicates that specify the desired computation
by the user. 
We will specify the CATARM query language in a formal grammar.

\subsection{Arabic linguistic computational model }
\label{s:design:lcm}

Several techniques~\cite{AEL07,Ham07,Abd07,MEl03} 
preprocessed the Arabic text and then used known text 
mining techniques that tend to Latin text and queries
to solve classification, indexing, named entity 
extraction and other problems.
We think that these techniques do not make use 
of Arabic language specific features.

We think that we can discover such features by building
a linguistic computational model of the Arabic language
based on its formal grammar and formal morphology
as defined in several Arabic linguistic 
sources~\cite{Sha73,Abd00,Abd001}.
We will build our morphological analyzer on top 
of this linguistic model in addition to the current existing
concatenative-based analysis techniques.

Let $L_q$ be the query language grammar, 
$L_A$ be the grammar that expresses the Arabic linguistic
computational model, and let $q$ be a user query.
We compute $\rho_q$, the grammar path that matches 
the query $q$ in $L_q$. 
We will explore methods that use $\rho_q$ to simplify 
$L_A$ before we start the morphological analysis. 
We think the possible simplifications that we can perform
here are Language specific as they will reflect the 
correspondence between the formal query language definition $L_q$, 
the usage patterns of $L_q$  by Arabic users, 
and $L_A$ the formal LCM for Arabic.

\subsection{Methodology}

\begin{figure}
\center{
\resizebox{.4\columnwidth}{!}
{\input{figs/hadith.pdftex_t}
\label{f:statemachine}
}
\caption{State-machine used to detect the chain of narrators}
}
\end{figure}

{\bf Case-based morphological analyzer.}
We experimented with several morphological analyzers.
We think that using finite state machines for morphological analysis
can be implemented in a more efficient way compared to other 
approaches. 
For example the concatenative rules used in Buckwalter~\cite{Tim04}
can be encoded in three FSM with linear execution time using 
the trie~\cite{datries} data structure.

The use of FSMs also suits our case-based approach. 
The controller can be a FSM machine on its own and its decision
will be another input into the internal morphological FSM structures. 
This also suits our choice of enriching the concatenative analysis
with a formal linguistic model that can also be encoded into FSMs. 
FSM can also be easily integrated with hidden Markov models (HMM)
extracted from the sets of documents the MA will consider.

The state machine in Firgure~\ref{f:statemachine} controls 
the concatenation based state machines of the morphological
analyzer.
The transitions are excited
by inputs from the morphological analyzer. 
A NAME label represents a proper name, 
an NRC label represents a narrator connector such as
\RL{`an} "on behalf of", \RL{qaal} "said", 
or one of their derivations. 
The IBN label corresponds to the word \RL{ibn} 
``the son of'' and represents a name connector.

The symbol $\tau_{\mbox{NMC}}$ is a threshold
that corresponds to the number of tolerated name connectors 
that may occur between two names. 
The symbol LIST$_{\mbox{NMC}}$ corresponds to the list 
of name connectors collected since the controller
started looping in the state NMC\_S. 
The symbol $\lambda_{\mbox{NMC}}$ is a parameter 
that corresponds to a relaxed tolerance measure that
the controller resorts to in case the words separating
two names were longer than $\tau_{\mbox{NMC}}$ but 
contained a name connector word such as IBN or NISBA.

The controller has four states that correspond to 
an abstract position in text relative to the next sanad. 
State TEXT\_S is the initial state and denotes that
the controller is outside the context of a sanad.
The controller moves to the state NAME\_S whenever
NAME is reported.
State NRC\_S means that the controller thinks it is in the context
of a narrator name after an NRC is reported.
State NMC\_S
indicates that the controller expects a name to appear within 
a tolerance threshold expressed by 
$\tau_{\mbox{NMC}}$ and $\lambda_{\mbox{NMC}}$.

Similarly, the NRC\_S state tolerates $\tau_{\mbox{NRC}}$ words 
before it gives up on its expectations.

We reach the NAME\_S state only when a
valid NAME is detected and we leave when no more NAME's 
are detected. 
The NRC\_S state can only be reached if an NRC is detected.
NMC\_S can only be reached from a NAME\_S state.

The controller reports a valid chain of narrators when a 
sequence of names
connected by narrator connectors appears. 
It marks the beginning of the hadith with the beginning of the 
current sequence,
and marks the end of the hadith with the beginning of the next 
sequence. 
It also marks the chain of narrators as the sequence itself. 
The definition of input labels such as NAME and IBN depends on 
the morphological analyzer. 
We readily implemented the above approach and confirmed its 
accuracy and efficiency compared to existing morphological
analyzers.

We will investigate enhancing our current case-based morphological
analyzer.
We will investigate methods to extend the internal structures 
of our analyzer to be able to represent a formal linguistic
computational model of the Arabic language.

{\bf Linguistic computational model of Arabic.}
We think that a linguistic computational model for the Arabic
language will perform better in applications where
relations between Arabic text documents are involved. 
Such a model helps capture features unique and specific 
to the Arabic language and make use of that in the analysis.

ElixirFM~\cite{Otakar:07} builds on the lexicon
of the Buckwalter analyzer and the annotations of the 
Prague Arabic dependency databank to build a formal 
computational model of the Arabic language.
The models is a set of formulas and rules. 
ElixirFM uses the computational model in a functional
morphology framework that uses Haskell and functional
programming. 
The analyzer works by generating morphological trees of 
decisions and then uses a disambiguation decision procedure 
to annotate the trees with contextual elements.

We will explore extracting our model from formal text book
definitions of the Arabic language~\cite{Abd00,Abd001} and seek
help from Arabic linguists to relax these formal rules to
allow common mistakes. 
We will encode the model in a grammar that we can synthesize
into an FSM and incorporate that with our current FSM 
morphological analyzer.

{\bf User query language. }
In Figure~\ref{f:statemachine} we showed an FSM 
that controls the morphological analysis and generates a
list of chains of narrators out of hadith text.
We think that it is overwhelming to ask a user who is not 
versed in FSM and in computational sciences in general 
to design such a controller.

The diagram in Figure~\ref{f:chain} depicts our vision of 
an equivalent query.
This query generates the list of chains by first passing the 
hadith text into the morphological analyzer which 
extracts names and their position in text. 
We iterate over the extracted names and compute the distance.
Whenever the distance is less than a threshold $\theta_1$
we add the word to a temporary chain of names. 
Otherwise, we execute a sequential block to 
{\bf 1.} tabulate the chain into a temporary table containing 
the list of chains. 
{\bf 2.} Create a new chain whith the last element.
Finally, we filter those chains whose cardinality 
is less than some threshold  $\theta_2$.

This query can be saved to a library of computational components
and invoked directly in another query 
as we saw in Figure~\ref{f:literature}.

\begin{figure}
\center{
\resizebox{.8\columnwidth}{!}
{\includegraphics{figs/detailed_submachine.pdf}
\label{f:chain}
}
\caption{CATARM detailed diagram for the chain of narrators extractor}
}
\end{figure}

The CATARM query language will be expressed in a grammar 
where the atomic elements are user defined constants, or
entities and tags extracted by the morphological analyzer. 
The language will define terms, expressions, and predicates
using several connectives that express set operations, 
relational operations, Boolean operations, scalar operations, 
statistical tools, text algorithms, and graph algorithms. 
We need to provide a front end tool to support this query
language.

{\bf Simplify LCM based on the query.}
Given
L$L_q$ the query language, 
$L_A$ the Arabic linguistic
computational model, and  $q$ a user query
we compute $\rho_q$, the grammar path that matches 
the query $q$ in $L_q$. 
We plan to inspect methods to overapproximate 
the morphemes and the transitions in $L_A$ 
that can be satisfied by morphemes and transitions
in $\rho_q$. 
This can simplify the number of states we need to 
look at under $L_A$.

This task is not trivial though, since $\rho_q$ may contain
states and transitions that are not always morphemes and 
transition between morphemes. 
We will consider a hierarchical methodology that follows 
the hierarchical structure of the Arabic language expressed
in $L_A$. For example,
we may consider sets of connected states in $\rho_q$ 
containing a single morpheme as a state in one hierarchy level 
in $L_A$.
We will perform this simplification before starting the 
morphological analysis since we already have to analyze
the query $q$ and compute $\rho_q$ before we start
the analysis.

{\bf Simplify LCM and MA based on set of documents.}
We will analyze the set of input documents given to a
morphological analyzer component and use the analysis
to further improve the morphological analysis. 
We will extract statistical features from these documents 
to help order the solutions considered by the analyzer
to consider the most probable first.

We will consider the use of HMM models 
built on the fly while processing the documents.
We will also consider the use of
latent semantic indexing~\cite{LSI89} techniques. 
We are considering these two techniques as they are 
both language independent and can both be computed and
adapted on the fly.

{\bf Entity and relational extractor.}
We will build a set of entity and relational extractors. 
The entities can be proper names, dates, locations,
and qualification adjectives. 
We will explore the use of pairing techniques to establish 
relations between the extracted entities.
The relations will form graphs as in the case studies presented
so far. The nodes represent the entities and the previously 
extracted relations and the edges represent the newly discovered 
relations.

The entity and relational extraction components 
may take as input a template for an expected relation and 
may return a graph that describes whether and where 
this relation exists between the subject documents.

We will need to augment the tags in the lexicons of the 
morphological analyzer to allow for better entity and relational
extractors. 
We plan to make use of existing annotated tree banks and of
modern Arabic dictionaries to do that.

{\bf Statistical components.}
We will make use and build statistical components that can be used
in CATARM. 
We will need statistical components to rank solutions, compute 
probable decisions, estimating good thresholds.

{\bf Graph algorithms.}
The relations extracted from the text documents will often 
be expressed in terms of graphs.
We will need textbook graph algorithms such as traversals, 
topological sorts, spanning trees, strongly connected 
components, minimum flow  and other algorithms.
For example, in the literature case study, the discovery
of articulation vertices may point to important narrators 
who may have an insignificant number of narrations 
but still connect narrator communities.

{\bf Query solver.}
This is a logical solver that will solve query expressions to \
resolve matching text and relations.
We will need to use existing off the shelf constraint solvers 
and set theory satisfiability solvers to address our queries. 
The solver should return and rank approximate matches.
That is if a three entities out of four required in a relation are
found, the solution is still worth reporting.

The solver will call the morphological analyzer to compute 
morphological matches to the atomic elements of the query. 
Then each operation in the query such as a set intersection
operation will correspond to a computational component. 
The solver will call these components and integrate their
results.

The solver may make use of existing propositional~\cite{MiniSAT} or
higher order logic~\cite{Z3} solvers if needed. 
These solvers can solve satisfiability questions.
We will explore the feasibility and utility 
of extracting a necessary set of 
constraints from the query and answering that with an off-the-shelf
logic solver.

{\bf Case studies.}
We will work on developing the necessary components to complete 
the solution for the hadith literature and the security case 
studies.
We will also work on the health and network traffic case studies.
While working on the case studies, we will add the components
we build to our library of components. 
This will enrich CATARM with a useful set of computational 
components.

{\bf Development methodology.}
We will integrate all the mentioned techniques in a scalable 
framework for Arabic text analysis.
The framework as a whole will allow for exhausting Arabic text 
analysis and allow for extensions of the framework.
A user should be able to iteratively and repetitively apply the 
abode techniques on his input text.

We will follow an eXtreme Programming 
methodology in the process of building CATARM.
We will use open source tools, libraries, and frameworks to 
build our libraries.
We will build on existing open source programming and text 
mining tools as well as open source Arabic text analysis tools 
based on NLP such as MADA+TOKAN~\cite{Rot08}, 
NEMLAR~\cite{RAl09}, and IJAES~\cite{Int09}.

@@

{\bf Visualization.} 
We will build a visualization tool that allows fast 
and guided inspection and correction mechanism. 
The CATARM visualization aid will be a tag and relation color
sensitive browser.
The interface will allow a fast and guided inspection and 
correction mechanism.

\subsection{Research team}

\subsection{Work plan}

\subsection{Budget}

@@

We anticipate that the research will lead to the production of 
CATARM, an Arabic text analysis framework that allows relational 
queries between Arabic text documents.
The components of the framework are listed and discussed
in details in Section~\ref{s:design}.

The following is a list of the anticipated components They are and are repeated for convenience as follows.
\begin{enumerate}
\item Case-based morphological analyzer 
\item Linguistic computational model for Arabic
\item User query language
\item Optimization methodologies
\begin{enumerate}
\item Use of language specific features 
\item Use of data in question specific features
\end{enumerate}
\item Entity and relational extractors
\item Statistical tools
\item Graph and text algorithms
\item Query solver
\item Case studies 
\begin{enumerate}
\item Hadith literature
\item Security
\item Health
\item Network content
\end{enumerate}
\end{enumerate}
